Untitled

### 2024-02-13 niceAi.Rmd
more screening

# Support functions and discrete Gaussian class construction 
```{r supportHMM, include=FALSE}
# order the states of the HMM to allow it to order them as low ret/mid/ret/high ret. 
label_switch <- function(mod,labels) {
  # labels is vector, first element is new label for original state 1, second is new label for original state 2, etc.
  if(!is(mod,"depmix") || !is(mod,"depmix.fitted")) stop("this function is for depmix models")
  n_states <- mod@nstates
  if(length(labels) != n_states || length(unique(labels)) != n_states || !(all(labels) %in% 1:n_states)) {
    stop("labels needs to be a vector of unique integers between 1 and", n_states)
  }
  inv_labels <- sapply(1:n_states,function(x) which(labels == x))
  tmp <- mod
  # relabel prior
  ppars <- getpars(mod@prior)
  fpars <- getpars(mod@prior,which="fixed")
  out_pars <- as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels]))
  out_fixed <- as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels]))
  if(!tmp@prior@family$link=="identity") tmp@prior@family$base <- labels[tmp@prior@family$base]
  # relabel transition
  for(i in 1:n_states) {
    ppars <- getpars(mod@transition[[inv_labels[i]]])
    fpars <- getpars(mod@transition[[inv_labels[i]]], which="fixed")
    out_pars <- c(out_pars,as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels])))
    out_fixed <- c(out_fixed,as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels])))
    tmp@transition[[i]] <- mod@transition[[inv_labels[i]]]
    if(!tmp@transition[[i]]@family$link=="identity") tmp@transition[[i]]@family$base <- labels[tmp@transition[[i]]@family$base]
    #out_pars <- c(out_pars,getpars(mod@transition[[inv_labels[i]]]))
  }
  # relabel response
  for(i in 1:n_states) {
    out_pars <- c(out_pars,unlist(lapply(mod@response[[inv_labels[i]]],getpars)))
    out_fixed <- c(out_fixed,unlist(lapply(mod@response[[inv_labels[i]]],getpars,which="fixed")))
  }
  tmp <- setpars(tmp,out_fixed,which="fixed")
  tmp <- setpars(tmp,out_pars)
  if(is(tmp,"depmix.fitted")) tmp@posterior <- viterbi(tmp)
  return(tmp)
}

# ORDERING STATES WHEN RESPONSE IS BINS
order_mod <- function(mod) {
  ns <- nstates(mod)
  sum <- rep(0.0,ns)
  for(i in 1:ns) {
    # Expected return in each state calculation (5 bins, for binned returns)
    sum[i] = sum(1:5*mod@response[[i]][[1]]@parameters$coefficients)
  }
  # reordering the states
  mod <- label_switch(mod, rank(sum))
  return(mod)
}

# ORDERING STATES WHEN RESPONSE IS GAUSSIAN
order_mod_gauss <- function(mod) {
  ns <- nstates(mod)
  sum <- rep(0.0,ns)
  for(i in 1:ns) {
    # Expected return in each state calculation 
    sum[i] = mod@response[[i]][[1]]@parameters$mu
  }
  # reordering the states
  mod <- label_switch(mod, rank(sum))
  return(mod)
}

@@

setMethod("discgaus", 
          signature(y="ANY"), 
          function(y,pstart=NULL,fixed=NULL, breaks = c(-Inf, 0:19 + .5, Inf), ...) {
            y <- matrix(y,length(y))
            x <- matrix(1)
            parameters <- list()
            npar <- 2
            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
            if(!is.null(pstart)) {
              if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
              parameters$mu <- pstart[1]
              parameters$sigma <- pstart[2]
            }
            mod <- new("discgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar, breaks=breaks)
            mod
          }
)

@@

setMethod("dens","truncdiscgaus",
          function(object,log=FALSE) {
            breaks <- c(object@min, object@breaks[object@breaks > object@min & object@breaks < object@max], object@max)
            p <- pnorm(breaks[-1], mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(breaks[-length(breaks)], mean = object@parameters$mu, sd = object@parameters$sigma)
            p <- p/sum(p)
            if(log) return(log(p[as.numeric(cut(object@y, breaks=object@breaks))])) else return(p[as.numeric(cut(object@y, breaks=object@breaks))])
          }
)

@@

setMethod("truncdiscgaus", 
          signature(y="ANY"), 
          function(y,pstart=NULL,fixed=NULL, breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20, ...) {
            y <- matrix(y,length(y))
            x <- matrix(1)
            parameters <- list()
            npar <- 2
            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
            if(!is.null(pstart)) {
              if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
              parameters$mu <- pstart[1]
              parameters$sigma <- pstart[2]
            }
            mod <- new("truncdiscgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar, breaks=breaks, min=min, max=max)
            mod
          }
)
```

```{r}
# Specify simple models to copy transition and prior
inv_simple <- list()
for(i in 2:6) {
  inv_simple[[i]] <- depmix(investment_bin ~ 1, data = origin_dat, nstates = i, transition = ~ invPnL, family=multinomial("mlogit"), ntimes = rep(10,381))
}

# Specify truncated gaussian models for 3 state model 
rModels3 <- list(
  list(
    truncdiscgaus(origin_dat$investment,pstart=c(5,2), min=-.5, max=20.5) # first guassian initialised with mean of 5 and sd of 2
  ),
  list(
    truncdiscgaus(origin_dat$investment,pstart=c(10,2), min=-.5, max=20.5)
  ),
  list(
    truncdiscgaus(origin_dat$investment,pstart=c(15,2), min=-.5, max=20.5)
  )
)

inv_dmod_3 <- makeDepmix(response=rModels3,transition=inv_simple[[3]]@transition,prior=inv_simple[[3]]@prior,ntimes = rep(10,381), homogeneous=FALSE)

inv_fdmod_3 <- fit(inv_dmod_3, emcontrol=em.control(random.start = FALSE))  #random.start = FALSE -> use parameters supplied as starting values

mod_3S <- order_mod_gauss(inv_fdmod_3)

@@

```{r, include=F}
#ORDER THE STATES OF THE HMM

state1 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[1]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))

state2 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[2]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))

state3 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[3]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))

investment <- seq(0:20) - 1

inv_resp_3S <- as.data.frame(cbind(investment,state1,state2,state3)) %>% 
  pivot_longer(cols=c("state1","state2","state3"),
                    names_to='investor_state',
                    values_to='probability')

@@

```{r}
inv_resp_3S %>% mutate( investor_state = fct_recode(investor_state, "low-trust" = "state1", "medium-trust" = "state2", "high-trust" = "state3"))%>%
        ggplot( aes(x = investment,
                    y = probability,
                    fill = investor_state)) +
  geom_bar(stat = "identity",
           position = "dodge") + 
  labs(fill='Latent investor state') +
  theme_bw() + 
  theme(legend.position = "bottom")

@@

```{r}
# Here pars takes the form of a list ns matrices, one for each state from 1 to ns.
plot_transitions_PnL <- function(fmod) {

@@

  trans_prob <- trans_prob %>% mutate(from=recode(from, 
                         `1`="low-trust",
                         `2`="medium-trust",
                         `3`="high-trust"),
                        to=recode(to, 
                         `1`="low-trust",
                         `2`="medium-trust",
                         `3`="high-trust")) %>% 
    mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
    mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))

#     iris %>%
#    mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>%
# ggplot() + 
#    geom_histogram(aes(Petal.Width))+ 
#    facet_grid(Species~.)
#   
  ggplot(trans_prob,aes(x=PnL,y=prob, colour = as.factor(to))) + geom_line() + facet_wrap(~from) + ylim(c(0,1)) +
      theme(legend.position = "bottom")
}
```

```{r}
plot_transitions_PnL(mod_3S)
```
```{r}
ns <- 3
pars <- NULL

@@

pars[[2]]["to_2","invPnL"]

@@

```{r}
plot_transitions_input <- function(pars_list) {

  ns <- nrow(pars_list[[1]])  # number of states

  for(from in 1:ns) {
    pars <- pars_list[[from]]
    for(to in 1:ns) {
      x <- trans_prob[trans_prob$from == from & trans_prob$to == to,"PnL"]
      y[,to] <- exp(pars[to,1] + pars[to,2]*x)
    }
    y <- y/rowSums(y)
    for(to in 1:ns) {
      trans_prob$prob[trans_prob$from == from & trans_prob$to == to] <- y[,to]
    }
  }

  trans_prob <- trans_prob %>% mutate(from=recode(from, 
                         `1`="low-trust",
                         `2`="medium-trust",
                         `3`="high-trust"),
                        to=recode(to, 
                         `1`="low-trust",
                         `2`="medium-trust",
                         `3`="high-trust")) %>% 
    mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
    mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))

  ggplot(trans_prob,aes(x=PnL,y=prob, colour = as.factor(to))) + geom_line() + facet_wrap(~from) + ylim(c(0,1))+
      theme(legend.position = "bottom")
}

```{r}
# pars has the format [[from]]["to","x"]
pars[[1]]["to_1","invPnL"] <- 0
pars[[1]]["to_1","intercept"] <-  10

pars[[2]]["to_1","intercept"] <- 3
#pars
plot_transitions_input(pars)
```

>>>>

# Support functions and discrete Gaussian class construction 
```{r supportHMM, include=FALSE}
# # order the states of the HMM to allow it to order them as low ret/mid/ret/high ret. 
# label_switch <- function(mod,labels) {
#   # labels is vector, first element is new label for original state 1, second is new label for original state 2, etc.
#   if(!is(mod,"depmix") || !is(mod,"depmix.fitted")) stop("this function is for depmix models")
#   n_states <- mod@nstates
#   if(length(labels) != n_states || length(unique(labels)) != n_states || !(all(labels) %in% 1:n_states)) {
#     stop("labels needs to be a vector of unique integers between 1 and", n_states)
#   }
#   inv_labels <- sapply(1:n_states,function(x) which(labels == x))
#   tmp <- mod
#   # relabel prior
#   ppars <- getpars(mod@prior)
#   fpars <- getpars(mod@prior,which="fixed")
#   out_pars <- as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels]))
#   out_fixed <- as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels]))
#   if(!tmp@prior@family$link=="identity") tmp@prior@family$base <- labels[tmp@prior@family$base]
#   # relabel transition
#   for(i in 1:n_states) {
#     ppars <- getpars(mod@transition[[inv_labels[i]]])
#     fpars <- getpars(mod@transition[[inv_labels[i]]], which="fixed")
#     out_pars <- c(out_pars,as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels])))
#     out_fixed <- c(out_fixed,as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels])))
#     tmp@transition[[i]] <- mod@transition[[inv_labels[i]]]
#     if(!tmp@transition[[i]]@family$link=="identity") tmp@transition[[i]]@family$base <- labels[tmp@transition[[i]]@family$base]
#     #out_pars <- c(out_pars,getpars(mod@transition[[inv_labels[i]]]))
#   }
#   # relabel response
#   for(i in 1:n_states) {
#     out_pars <- c(out_pars,unlist(lapply(mod@response[[inv_labels[i]]],getpars)))
#     out_fixed <- c(out_fixed,unlist(lapply(mod@response[[inv_labels[i]]],getpars,which="fixed")))
#   }
#   tmp <- setpars(tmp,out_fixed,which="fixed")
#   tmp <- setpars(tmp,out_pars)
#   if(is(tmp,"depmix.fitted")) tmp@posterior <- viterbi(tmp)
#   return(tmp)
# }
# 
# # ORDERING STATES WHEN RESPONSE IS BINS
# order_mod <- function(mod) {
#   ns <- nstates(mod)
#   sum <- rep(0.0,ns)
#   for(i in 1:ns) {
#     # Expected return in each state calculation (5 bins, for binned returns)
#     sum[i] = sum(1:5*mod@response[[i]][[1]]@parameters$coefficients)
#   }
#   # reordering the states
#   mod <- label_switch(mod, rank(sum))
#   return(mod)
# }
# 
# # ORDERING STATES WHEN RESPONSE IS GAUSSIAN
# order_mod_gauss <- function(mod) {
#   ns <- nstates(mod)
#   sum <- rep(0.0,ns)
#   for(i in 1:ns) {
#     # Expected return in each state calculation 
#     sum[i] = mod@response[[i]][[1]]@parameters$mu
#   }
#   # reordering the states
#   mod <- label_switch(mod, rank(sum))
#   return(mod)
# }

@@

setMethod("discgaus", 
          signature(y="ANY"), 
          function(y,pstart=NULL,fixed=NULL, breaks = c(-Inf, seq(0, 19) + .5, Inf), ...) {
            y <- matrix(y,length(y))
            x <- matrix(1)
            parameters <- list()
            npar <- 2
            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
            if(!is.null(pstart)) {
              if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
              parameters$mu <- pstart[1]
              parameters$sigma <- pstart[2]
            } else {
              parameters <- list(mu = 10, sigma = 3)
            }
            mod <- new("discgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar, breaks=breaks)
            mod
          }
)

@@

setMethod("dens","truncdiscgaus",
          function(object,log=FALSE) {
            breaks <- c(object@min, object@breaks[object@breaks > object@min & object@breaks < object@max], object@max)
            prec <- pnorm(object@max, mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(object@min, mean = object@parameters$mu, sd = object@parameters$sigma)
            if(prec < 1e-12) {
              # return uniform probabilities as precision is likely off
              p <- rep(1/(length(breaks)-1), length(object@y))
              if(log) return(log(p)) else return(p)
            } else {
              p <- pnorm(breaks[-1], mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(breaks[-length(breaks)], mean = object@parameters$mu, sd = object@parameters$sigma)
              p <- p/sum(p)
              if(log) return(log(p[as.numeric(cut(object@y, breaks=object@breaks))])) else return(p[as.numeric(cut(object@y, breaks=object@breaks))])
            }
          }
)

@@

setMethod("truncdiscgaus", 
          signature(y="ANY"), 
          function(y,pstart=NULL,fixed=NULL, breaks = c(-Inf, seq(0, 19) + .5, Inf), min=-0.5, max=20.5, ...) {
            y <- matrix(y,length(y))
            x <- matrix(1)
            parameters <- list()
            npar <- 2
            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
            if(!is.null(pstart)) {
              if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
              parameters$mu <- pstart[1]
              parameters$sigma <- pstart[2]
            }
            mod <- new("truncdiscgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar, breaks=breaks, min=min, max=max)
            mod
          }
)

# define a response class which only contains the standard slots, no additional slots
setClass("vtdgaus", contains="response", slots=c(yield="numeric"))

setGeneric("vtdgaus", function(y, pstart = NULL, fixed = NULL, ...) standardGeneric("vtdgaus"))

setMethod("vtdgaus", 
          signature(y="ANY"), 
          function(y, yield, pstart=NULL,fixed=NULL, ...) {
            y <- matrix(y,length(y))
            x <- matrix(1)
            parameters <- list()
            npar <- 2
            if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
            if(!is.null(pstart)) {
              if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
              parameters$mu <- pstart[1]
              parameters$sigma <- pstart[2]
            } else {
              parameters <- list(mu=.5,sigma=1)
            }
            mod <- new("vtdgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar, yield=yield)
            mod
          }
)

setMethod("show","vtdgaus",
  function(object) {
    cat("Gaussian with variable discrete support for percentage responses\n")
    cat("Parameters: \n")
    cat("mu: ", object@parameters$mu, "\n")
    cat("sigma: ", object@parameters$sigma, "\n")
  }
)

setMethod("dens","vtdgaus",
  function(object,log=FALSE) {
      # determine cuts for pnorm based on 
      #p <- pnorm(sapply(object@y + .5*(1/object@yield), function(x) min(x,1.001)), mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(sapply(object@y - .5*(1/object@yield),function(x) max(x,-0.001)), mean = object@parameters$mu, sd = object@parameters$sigma)
      #norm <- (pnorm(1 + .001, mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(-0.001, mean = object@parameters$mu, sd = object@parameters$sigma))
      #p <- p/norm

      # alternatively, normalize factor based on size of cuts

      prec <- pnorm(1 + .5*(1/60), mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(0 - .5*(1/60), mean = object@parameters$mu, sd = object@parameters$sigma)
      # check for issues with precision, using highest yield
      if(prec < 1e-12) {
        p <- 1/(object@yield + 1)
      } else {
        p <- pnorm(object@y + .5*(1/object@yield), mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(object@y - .5*(1/object@yield), mean = object@parameters$mu, sd = object@parameters$sigma)
        norm <- pnorm(1 + .5*(1/object@yield), mean = object@parameters$mu, sd = object@parameters$sigma) - pnorm(0 - .5*(1/object@yield), mean = object@parameters$mu, sd = object@parameters$sigma)
        p <- p/norm
      }
      # probability when yield == 0 is always 1 
      p[object@yield == 0] <- 1
      if(log) return(log(p)) else return(p)
    }
)

setMethod("setpars","vtdgaus",
          function(object, values, which="pars", ...) {
            npar <- npar(object)
            if(length(values)!=npar) stop("length of 'values' must be",npar)
            # determine whether parameters or fixed constraints are being set
            nms <- names(object@parameters)
            switch(which,
                   "pars"= {
                     object@parameters$mu <- values[1]
                     object@parameters$sigma <- values[2]
                   },
                   "fixed" = {
                     object@fixed <- as.logical(values)
                   }
            )
            names(object@parameters) <- nms
            return(object)
          }
)

setMethod("getpars","vtdgaus",
          function(object,which="pars",...) {
            switch(which,
                   "pars" = {
                     parameters <- numeric()
                     parameters <- unlist(object@parameters)
                     pars <- parameters
                   },
                   "fixed" = {
                     pars <- object@fixed
                   }
            )
            return(pars)
          }
)

setMethod("fit","vtdgaus",
          function(object,w) {
            if(missing(w)) w <- NULL
            if(!is.null(w)) {
              negLL <- function(pars) {
                object <- setpars(object, c(pars[1], exp(pars[2])))
                -sum(w*log(dens(object)))
              }
            } else {
              negLL <- function(pars) {
                object <- setpars(object, c(pars[1], exp(pars[2])))
                -sum(log(dens(object)))
              }
            }
            pars <- optim(c(object@parameters$mu, log(object@parameters$sigma)), fn=negLL)$par
            object <- setpars(object,c(pars[1], exp(pars[2])))
            object
          }
)
```

```{r}
#| label: HMM-support-functions
#| include: false
# order the states of the HMM to allow it to order them as low ret/mid/ret/high ret. 
label_switch <- function(mod,labels) {
  # labels is vector, first element is new label for original state 1, second is new label for original state 2, etc.
  if(!is(mod,"depmix") || !is(mod,"depmix.fitted")) stop("this function is for depmix models")
  n_states <- mod@nstates
  if(length(labels) != n_states || length(unique(labels)) != n_states || !(all(labels) %in% 1:n_states)) {
    stop("labels needs to be a vector of unique integers between 1 and", n_states)
  }
  inv_labels <- sapply(1:n_states, function(x) which(labels == x))
  tmp <- mod
  # relabel prior
  ppars <- getpars(mod@prior)
  fpars <- getpars(mod@prior, which="fixed")
  out_pars <- as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels]))
  out_fixed <- as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels]))
  if(!tmp@prior@family$link=="identity") tmp@prior@family$base <- labels[tmp@prior@family$base]
  # relabel transition
  for(i in 1:n_states) {
    ppars <- getpars(mod@transition[[inv_labels[i]]])
    fpars <- getpars(mod@transition[[inv_labels[i]]], which="fixed")
    out_pars <- c(out_pars,as.numeric(t(matrix(ppars,nrow=length(ppars)/n_states,byrow = TRUE)[,inv_labels])))
    out_fixed <- c(out_fixed,as.logical(t(matrix(fpars,nrow=length(fpars)/n_states,byrow = TRUE)[,inv_labels])))
    tmp@transition[[i]] <- mod@transition[[inv_labels[i]]]
    if(!tmp@transition[[i]]@family$link=="identity") tmp@transition[[i]]@family$base <- labels[tmp@transition[[i]]@family$base]
    #out_pars <- c(out_pars,getpars(mod@transition[[inv_labels[i]]]))
  }
  # relabel response
  for(i in 1:n_states) {
    out_pars <- c(out_pars,unlist(lapply(mod@response[[inv_labels[i]]],getpars)))
    out_fixed <- c(out_fixed,unlist(lapply(mod@response[[inv_labels[i]]],getpars,which="fixed")))
  }
  tmp <- setpars(tmp,out_fixed,which="fixed")
  tmp <- setpars(tmp,out_pars)
  if(is(tmp,"depmix.fitted")) tmp@posterior <- viterbi(tmp)
  return(tmp)
}

# ordering states for a truncated discretised Gaussian
order_mod_truncdiscgaus <- function(mod) {
  ns <- nstates(mod)
  sum <- rep(0.0,ns)
  for(i in 1:ns) {
    # Expected return in each state calculation
    tpars <- getpars(mod@response[[i]][[1]])
    dmod <- truncdiscgaus(seq(0, 20), pstart = tpars, min=-0.5, max=20.5)
    sum[i] <- sum(seq(0,20)*dens(dmod))
  }
  cat("Expected values for states: ",sum, "\n")
  # reordering the states
  mod <- label_switch(mod, rank(sum, ties.method = "first"))
  return(mod)
}

# ordering states for a variable response truncated discretised Gaussian
order_mod_vtdgaus <- function(mod) {
  ns <- nstates(mod)
  sum <- rep(0.0,ns)
  for(i in 1:ns) {
    # Expected return in each state calculation
    tpars <- getpars(mod@response[[i]][[1]])
    dmod <- vtdgaus(seq(0, 1, length=61), pstart = tpars, yield=rep(60, 61))
    sum[i] <- sum(seq(0, 1, length=61)*dens(dmod))
  }
  # reordering the states
  mod <- label_switch(mod, rank(sum, ties.method = "first"))
  return(mod)
}
```

```{r}
#| label: fit-HMM-investor-models
#| include: false
#| cache: true

inv_fdmod <- list()

set.seed(20230824)

for(i in 2:4) {
  inv_simple <- depmix(investment_bin ~ 1, data = origin_dat, nstates = i, transition = ~ invPnL, family=multinomial("mlogit"), ntimes = rep(10,381))

  mus <- (1:i)*(20/(i+1))
  sigmas <- rep((20/(i+1)), i)
  rModels_inv <- list()
  for(s in 1:i) {
    rModels_inv[[s]] <- list(truncdiscgaus(origin_dat$investment, pstart=c(mus[s],sigmas[s]), min=-0.5, max=20.5))
  }

  #rModels_inv <- rep(list(list(truncdiscgaus(origin_dat$investment, pstart=c(10,5), min=-0.5, max=20.5))), each=i)

  inv_dmod <- makeDepmix(response=rModels_inv, transition=inv_simple@transition, prior=inv_simple@prior, ntimes = rep(10,381), homogeneous=FALSE)

  inv_fdmod[[i]] <- fit(inv_dmod, emcontrol = em.control(random.start = FALSE))

  #inv_fdmod[[i]] <- multistart(inv_dmod, nstart=20)
}

#ORDER THE STATES OF THE HMM
mod_3S  <- order_mod_truncdiscgaus(inv_fdmod[[3]])

@@

```{r}
# # Specify simple models to copy transition and prior
# inv_simple <- list()
# for(i in 2:6) {
#   inv_simple[[i]] <- depmix(investment_bin ~ 1, data = origin_dat, nstates = i, transition = ~ invPnL, family=multinomial("mlogit"), ntimes = rep(10,381))
# }
# 
# 
# # Specify truncated gaussian models for 3 state model 
# rModels3 <- list(
#   list(
#     truncdiscgaus(origin_dat$investment,pstart=c(5,2), min=-.5, max=20.5) # first guassian initialised with mean of 5 and sd of 2
#   ),
#   list(
#     truncdiscgaus(origin_dat$investment,pstart=c(10,2), min=-.5, max=20.5)
#   ),
#   list(
#     truncdiscgaus(origin_dat$investment,pstart=c(15,2), min=-.5, max=20.5)
#   )
# )
# 
# inv_dmod_3 <- makeDepmix(response=rModels3,transition=inv_simple[[3]]@transition,prior=inv_simple[[3]]@prior,ntimes = rep(10,381), homogeneous=FALSE)
# 
# inv_fdmod_3 <- fit(inv_dmod_3, emcontrol=em.control(random.start = FALSE))  #random.start = FALSE -> use parameters supplied as starting values
# 
# mod_3S <- order_mod_gauss(inv_fdmod_3)

@@

```{r, include=F}

# 
# state1 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[1]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))
# 
# state2 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[2]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))
# 
# state3 <- dens(truncdiscgaus(seq(0:20) - 1,pstart=c(unlist(mod_3S@response[[3]][[1]]@parameters)), breaks = c(-Inf, 0:19 + .5, Inf), min=0, max=20))
# 
# investment <- seq(0:20) - 1
# 
# inv_resp_3S <- as.data.frame(cbind(investment,state1,state2,state3)) %>% 
#   pivot_longer(cols=c("state1","state2","state3"),
#                     names_to='investor_state',
#                     values_to='probability')

#####################
inv_policy <- list()
for(i in 1:3) {
  inv_policy[[i]] <- dens(truncdiscgaus(seq(0,20),  pstart=c(unlist(mod_3S@response[[i]][[1]]@parameters)), breaks = c(-Inf, seq(0, 19) + .5, Inf), min=-0.5, max=20.5))
}

investment <- seq(0,20)

df_inv_policy <- data.frame(investor_state = rep(1:3, each=length(investment)),
                            investment,
                            probability = unlist(inv_policy))

#####################

```{r}
#| label: fig-policy-plot-3-state-HMM
#| echo: false
#| fig.cap: Probability mass function of the investor's policy conditional on its latent state as an output of the best fitting 3 state HMM using a truncated discretised Gaussian as a response function
#| fig.align: center
#| fig.width: 8
#| fig.height: 4

plotinvHMM <- ggplot(df_inv_policy ,                            
       aes(x = investment,
           y = probability,
           fill = factor(paste0("s",investor_state)))) +
  geom_bar(stat = "identity",
           position = "dodge") + 
  labs(fill='Latent investor state') +
  theme_minimal() + 
  theme(legend.position = "bottom")

plotinvHMM

```{r}
# inv_resp_3S %>% mutate( investor_state = fct_recode(investor_state, "low-trust" = "state1", "medium-trust" = "state2", "high-trust" = "state3"))%>%
#         ggplot( aes(x = investment,
#                     y = probability,
#                     fill = investor_state)) +
#   geom_bar(stat = "identity",
#            position = "dodge") + 
#   labs(fill='Latent investor state') +
#   theme_bw() + 
#   theme(legend.position = "bottom")

@@

```{r}
#| label: transition-plot-function
#| include: false
#| warning: false
plot_transitions_PnL <- function(fmod) {

@@

  trans_prob <- trans_prob %>%
    mutate(from = paste0("s", from),
           to = paste0("s", to))

  # trans_prob <- trans_prob %>% mutate(from=recode(from, 
  #                        `1`="low-trust",
  #                        `2`="medium-trust",
  #                        `3`="high-trust"),
  #                       to=recode(to, 
  #                        `1`="low-trust",
  #                        `2`="medium-trust",
  #                        `3`="high-trust")) %>% 
  #   mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
  #   mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))
  ggplot(trans_prob,aes(x=PnL,y=prob, colour = as.factor(to))) + geom_line() + facet_wrap(~from) + ylim(c(0,1))
}
```

```{r}
#| label: transition-plot-investor-HMM-3-state
#| echo: false
#| fig.cap: Transition probabilities of the investor state as a function of investor gain in the current round. Each panel is the state transitioned from, and each line shows the probability of transitioning to the state identified by the line color.
#| fig.align: center
#| fig.width: 8
#| fig.height: 4
plot_transitions_PnL(mod_3S) + labs(x = "Investor gain", y = "Probability", color='Transition to') + theme_bw() + theme(legend.position = "bottom") 
```

```{r}
# plot_transitions_input <- function(pars_list) {
# 
#   ns <- nrow(pars_list[[1]])  # number of states
#   
#   trans_prob <- data.frame(
#     from = rep(1:ns, each=80*ns),
#     to = rep(1:ns, each=80),
#     PnL = seq(-20,60,length=80),
#     prob = 0
#   )
# 
#   y <- matrix(0.0,ncol=ns, nrow=80)
#   
#   for(from in 1:ns) {
#     pars <- pars_list[[from]]
#     for(to in 1:ns) {
#       x <- trans_prob[trans_prob$from == from & trans_prob$to == to,"PnL"]
#       y[,to] <- exp(pars[to,1] + pars[to,2]*x)
#     }
#     y <- y/rowSums(y)
#     for(to in 1:ns) {
#       trans_prob$prob[trans_prob$from == from & trans_prob$to == to] <- y[,to]
#     }
#   }
#   
#   trans_prob <- trans_prob %>% mutate(from=recode(from, 
#                          `1`="low-trust",
#                          `2`="medium-trust",
#                          `3`="high-trust"),
#                         to=recode(to, 
#                          `1`="low-trust",
#                          `2`="medium-trust",
#                          `3`="high-trust")) %>% 
#     mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
#     mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))
#    
#   ggplot(trans_prob,aes(x=PnL,y=prob, colour = as.factor(to))) + geom_line() + facet_wrap(~from) + ylim(c(0,1))+
#       theme(legend.position = "bottom")
# }
#

```{r}
plot_HMM_transitions <- function(ns, pars_mat) {

  trans_prob <- data.frame(
    from = rep(1:ns, each=100*ns),
    to = rep(1:ns, each=100),
    ret = seq(-20,60,length=100),
    probs = 0
  )

  y <- matrix(0.0,ncol=ns, nrow=100)

  for(from in 1:ns) {
  pars <- matrix(pars_mat[[from]], ncol=2)
  # print(pars)

    for(to in 1:ns) {
        x <- trans_prob[trans_prob$from == from & trans_prob$to == to,"ret"]
        y[,to] <- exp(pars[to,1] + pars[to,2]*x)
    }
    y <- y/rowSums(y)

    for(to in 1:ns) {
      trans_prob$probs[trans_prob$from == from & trans_prob$to == to] <- y[,to]
    }
  }

  df <- as.data.frame(trans_prob) %>% 
    mutate(from = recode(from, "1" = "low-trust", "2" = "medium-trust", "3" = "high-trust"),
           to = recode(to, "1" = "low-trust", "2" = "medium-trust", "3" = "high-trust") ) %>% 
    mutate(across(from, factor, levels=c("low-trust","medium-trust","high-trust"))) %>% 
    mutate(across(to, factor, levels=c("low-trust","medium-trust","high-trust")))

    # Create a separate data frame with the background colors
  bg_colors <- data.frame(
    from = factor(c("low-trust", "medium-trust", "high-trust"), levels=c("low-trust","medium-trust","high-trust"))
  )

  # plotting code...
  ggplot() +
    geom_rect(data = bg_colors, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf, fill = from), alpha = 0.1) +
    geom_line(data = df, aes(x = ret, y = probs, colour = as.factor(to))) +
    facet_wrap(~from, labeller = labeller(from = function(x) paste("From", x, "state on trial t"))) +
    ylim(c(0,1)) +
    scale_fill_manual(values = c("low-trust" = "red", "medium-trust" = "green", "high-trust" = "blue")) +
    labs(x = "Investor's net return on trial t", y = "Transition probability to \nState on trial t+1", color = 'State transitioned to') +
    theme_bw() +
    theme(legend.position = "bottom", 
          axis.title.y = element_text(size = 8, angle = 90, vjust = 1.5))

```{r}
ns <- 3
pars <- NULL

@@

# pars has the format [[from]]["to","x"]
pars[[2]]

plot_HMM_transitions (3,pars)

@@

```{r}
# pars has the format [[from]]["to","x"]
new_pars <- pars
new_pars[[1]]["to_1","invPnL"] <- 0
new_pars[[1]]["to_1","intercept"] <-  -10
new_pars[[1]]["to_3","intercept"] <-  -10

#new_pars[[2]]["to_1","intercept"] <- 5
plot_HMM_transitions (3,new_pars)
```
